A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Primer empujón variables

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature and humidity. In
  this research, we address this question by means of a spatial analysis
  of the incidence of COVID-19 in the provinces in Spain. Use of a spatial
  Seemingly Unrelated Regressions (SUR) approach allows us to model the
  incidence of reported cases of the disease per 100,000 population, as a
  function of temperature and humidity, while controlling for GDP per
  capita, population density, percentage of older adults in the
  population, and presence of mass transit systems. An interesting aspect
  of the spatial SUR approach is that it models incidence as a contagion
  process. Our results indicate that incidence of the disease is lower at
  higher temperatures. The evidence with respect to humidity is more
  mixed, with coefficients for this variable that are significant in only
  some equations. Our control variables also yield interesting insights,
  as population density and percentage of older adults display negative
  associations with incidence of COVID-19, whereas the presence of mass
  transit systems in the province is associated with a greater incidence
  of the disease.\\
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, whereas de Ángel Solá et al.~(2020) see less risk in the Caribean
Basin, Coelho et al.~(Coelho et al., 2020) warn that at least during the
exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. To further complicate matters, much of the relevant work
has yet to be peer-reviewed and therefore is open to change (see for
example the challenge of Harbert et al. (2020) to Araujo and Naimi
(2020)). In the United States, the National Academy of Sciences,
Engineering, and Medicine was engaged by the Office of the Executive for
guidance on this matter (see National Academies of Sciences, Engineering
and Medicine, 2020). Their conclusion summarizes the situation well (see
p.~6): ``Some limited data support a potential waning of cases in warmer
and more humid seasons, yet none are without major
limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic unfolds
could shed more light on the effects of climate on transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature and
humidity. Hence, here we report results from a spatial analysis of
incidence of COVID-19 in fifty provinces in Spain. Spain is one of the
countries hardest hit by the virus, and enacted lockdown measures on
March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
We combine data on reported cases of the disease with metereological
information, to create a spatio-temporal dataset covering a period of 30
days. We then join this dataset with provincial-level economic and
demographic information to act as controls to our key environmental
variables. These data are analyzed using a spatial Seemingly Unrelated
Regressions (SUR) approach, which allows us to model incidence of
COVID-19 as a spatial contagion process.

The results provide evidence of the effect of temperature on the
incidence of COVID-19. The evidence concerning humidity is more mixed:
while the direction of the effect is negative, as anticipated, the
coefficients for this variable are only significant in some of equations
in the system. Our control variables also provide some intriguing
insights. The results of this analysis provide support to the hypothesis
of seasonality of the novel SARS-CoV2, and should be of interest to
public health officials and policy makers wrestling with the question of
the trajectory of the pandemic.

@@

Explain the rationale for selecting the variables.

For example: The literature about COVID-19 suggested that population
density is the one of the most important proliferate cause of these
viscous, however this ill spread with different intensity at big cities
of the world. Controlling for socioeconomic characteristics the
objective of this paper is observe the effect of clime on COVID-19
proliferation.

Some questions are important to point out. Is not evident present a
diary association between the temprature (resp. humidity) with declarate
case. There is evidences (ref) that a time lag is necessary between the
day that the individuo inffect and the case is diagnosticate

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\);
\(X_t=(X^1,...,X^{k_t})\) is a \(N \times k_t\) matrix of the \(k_t\)
independent variables, \(X_i=(X^i_{st})\);
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

@@

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

\begin{equation}
y = X \beta + \epsilon
\end{equation}

As is the case with cross-sectional data, it is possible to test the
residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and
several tests have been developed to test the null hypothesis of spatial
independence (see López et al., 2014). When the null hypothesis is
rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. Spatial analytical
approaches were used to understand contagion-difussion processes in the
case of infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

@@

\begin{equation}
\label{eq:sur-slm}
\bf{A}y = X \beta + \epsilon \\
\epsilon =N(0,\Omega)
\end{equation}

\noindent where \(A =I_{TN}-\bf{\Lambda} \otimes W\) with
\(\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)\).

This specification assumes that incidence in a province (\(y_st\)) at
time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
of incidence in neighbouring provinces. Coefficients for the spatially
lagged variable are estimated for each time period \(\lambda_t\) and
identifies the intensity and the sign of the impacts of neighbourhood.
It is possible test the null hypotheis of identtical levels of spatial
dependence (\(\lambda_i=\lambda_j, \forall i,j\)). The correspond Wald
test is available in the \texttt{R} package \texttt{spsur}.

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
alternative methodologies could be use. By example, a dynamic spatial
panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
2014, Cap. 4), but those models are more dont take account correlation
between errors. Therefore, a spatial SUR approach is more reasonable for
our purpose.

@@

@@

@@

@@

>>>>

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature, humidity, and
  hours of sunshine. In this research, we address this question by means
  of a spatial analysis of the incidence of COVID-19 in the provinces in
  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
  allows us to model the incidence of reported cases of the disease per
  100,000 population, as a function of temperature and humidity, while
  controlling for GDP per capita, population density, percentage of older
  adults in the population, and presence of mass transit systems. An
  interesting aspect of the spatial SUR approach is that it models
  incidence as a contagion process. Our results indicate that incidence of
  the disease is lower at higher temperatures and higher levels of
  humidity, although coefficients for this variable are significant only
  in some equations. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.\\
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~({\textbf{???}}), examined data from China and
came to the conclusion that neither temperature nor ultraviolet indices
had an association with transmission of COVID-19. This is despite
previous research that has linked less exposure to UVB radiation to
higher prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature,
humidity, and sunshine. We adopt a population health approach, and
report results from a spatial model of the incidence of COVID-19 in
fifty provinces in Spain, one of the countries hardest hit by the
pandemic. We combine data on reported cases of the disease with
metereological information, to create a spatio-temporal dataset covering
a period of 30 days. We then join this dataset with provincial-level
economic and demographic information to act as controls to our key
environmental variables. These data are analyzed using a spatial
Seemingly Unrelated Regressions (SUR) approach, which allows us to model
incidence of COVID-19 as a spatial contagion process.

The results provide evidence of the effect of temperature, humidity, and
sunshine on the incidence of COVID-19. The clearest result with respect
to these variables is a lower incidence of COVID-19 at higher
temperatures and levels of humidity, while the opposite happens with
respect to hours of sunshine. Our control variables also provide some
intriguing insights. Higher incidence is associated with higher GDP per
capita and presence of mass transit systems in the province; in
contrast, population density and percentage of older adults display
negative associations with incidence of COVID-19. The results of this
analysis provide support to the hypothesis of seasonality of the novel
SARS-CoV2, and should be of interest to public health officials and
policy makers grappling with the question of the trajectory of the
pandemic.

@@

Two set of variables has been selected to explain the diary incidence
ratio in each Spanish province. The first, are climatic variables
(temperature, humidity and sunshine) and its impact has been argued in
the introduction section. The second, is a set of specific
characteristics of each province. The objective is include in our model
some control variables (socioeconomic characteristics) to capture the
probable heterogeneity of each spatial unit. All those variables are
constant along the temporal period analized in this research. In this
subsetion we present some evidences of the expected relationship between
the four control variables considered an the incidence ratio.

The structure of population is a key variable on the incidence. Several
papers related the relationship between a fast trasnmision of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC) and therefore a positive sign will be expected. Note that our
analysis is at provincial level - not a urban level - and therefore is
necessary to rethinking the relationship. In Spain the population is
concentrate (with som exception) in the mediterranean coast (Goerlich
and Cantarino (2013)), in general with high temperature and high
humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related
with the population structure include like control variable. Unlike the
population density the older people in Spain is concentrate in the core
of the Iberian Peninsula (see figure \ref{plotfer1}).

The per capita GDP also can generate differences between provinces. No
direct relationship can be observed between the GDP and the ratio of
incidence but the GDP is a proxy of inobservable information. This
varaible has been included in similar papers (Wang et al. (2020)). For
low DGPpc is more difficult maintain social distance (sing +); High
DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
difícl mantener el confinamiento.

Posibles relaciones causales Alto GDP -\textgreater{} mas movimiento de
la población. Se recibió más gente de otros sitios que podría venir con
el virus puesto! Alto GDP -\textgreater{} La actividad económica de la
industria y la construcción solo se paralizó unos días. Aquellas
provincias con alto GDP tienen mas astividad industrial/contrucción
-\textgreater{} más movimimiento de la población. Bajo GDP
-\textgreater{} La población con menores ingresos no puede ser tan
obediente a la orden de confinamiento (suele ser menos disciplinada).

All those variables will we include in log to capture non-linar
realtionship.

@@

\noindent where \(y_{t}=(y_{1t},...,y_{Nt})\) is a \(N \times 1\)
vector, and in our case \(y_{st}\) is the incidence ratio in the
province \(s\) (\(s=1,...,N\)) the day \(t\) \((t=1,...,T)\); \(X_t\) is
a \(N \times k_t\) matrix of the \(k_t\) independent variables, with
associated vector of coefficients \(\beta_t\),;
\(\beta_t=(\beta_{1t},...,\beta_{Nt})\) is a vector of coefficients and
\(\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})\) is the vector of
residuals.

@@

Equation (\ref{eq:sur-sim}) can be rewriten in compact form:

\begin{equation}
\bf{Y} = \bf{X} \beta + \epsilon
\label{eq:sur-sim-block}
\end{equation}

where \(\bf{Y}\) is now a vector of dimension \(NT \times 1\),
\(\bf{X}\) is a block-diagonal matrix \(NT \times K\) (with
\(K = \sum_t{k_t}\)) and \(\epsilon\) is a \(NT \times 1\) vector. Using
the Kronecker product notation (\(\otimes\)), the error matrix structure
is expressed concisely as:

\begin{equation}
E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
\end{equation}

As is the case with cross-sectional data, it is possible to test the
residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation,
and several tests have been developed to test the null hypothesis of
spatial independence (see López et al., 2014). When the null hypothesis
is rejected, several alternative specifications have been proposed to
include spatial effects (Anselin, 1988, see also 2016). In this paper we
consider a spatial SUR model that incorporates a spatial lag of the
dependent variable as an explanatory factor. Spatial analytical
approaches were used to understand contagion-difussion processes in the
case of infectious disease in general (e.g., Cliff et al., 1998) and the
2003-2004 SARS outbreak in particular (e.g., Meng et al., 2005; Cao et
al., 2010). While we are mindful of the same caveat that the novel
SARS-CoV2 may not follow the patterns of previous diseases, there is
still evidence from the United States that COVID-19 displays spatial
patterns that are consistent with a diffusion process (Desjardins et
al., 2020). For this reason, the spatial lag model is appropriate to
model incidence of COVID-19 geographically, since it accounts for
potential spatial patterns that result from a process of contagion, as
explained next.

@@

\begin{equation}
\label{eq:sur-slm}
\begin{aligned}
\bf{AY} = \bf{X} \beta + \epsilon \\
\epsilon =N(0,\bf{\Sigma})
\end{aligned}
\end{equation}

\noindent where \(\bf{A} =I_{TN}-\bf{\Gamma} \otimes W\) with
\(\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)\).

This specification assumes that incidence in a province (\(y_{st}\)) at
time \(t\) is partially determined by the weighted average (\(Wy_{st}\))
of incidence in neighbouring provinces. Coefficients for the spatially
lagged variable are estimated for each time period \(\rho_t\) and
identifies the intensity and the sign of the impacts of neighbourhood.
It is possible test the null hypothesis of identtical levels of spatial
dependence (\(\rho_i=\rho_j, \forall i,j\)). The correspond Wald test is
available in the \texttt{R} package \texttt{spsur}.

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
alternative methodologies could be use. By example, a dynamic spatial
panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
2014, Cap. 4), but those models don't take account correlation between
errors. Therefore, a spatial SUR approach is more reasonable for our
purpose.

@@

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plotfer1-1.pdf}
\caption{\label{Fig-quantile} Spatial distribution of Spanish density
population}
\end{figure}

@@

Limitaciones

\begin{itemize}
\item
  La temperatura puedes ser un factor pero cabría esperar que no tuviera
  un impacto lineal. Por el contrario deberiamos esperar un `punto de
  corte': Una temperatua mantenida superior a X grados durante siete
  días sea la mejor forma de incorporarla al modelo. En nuestro modelo
  (log-log) en incremento en un 1\% de la temperatura se asocia con un
  incremento beta\% en la incidencia. Esto es lo mismo si la temperatura
  es baja que si es alta.
\item
  Idem para la humedad y horas de sol
\item
  Los datos son `provisionales'. Hay gran confusión sobre la incidencia
  real. La ausencia de test de diagnóstico PCRs al inicio de la pandemia
  (también ahora) puede desvirtuar el número de casos dianosticados.
\item
  Los datos oficiales (que tampoco son fiables) son reportados a nivel
  de Comunidades Autónomas. La recopilación de datos a nivel de
  provincial son el resultado de un esfuerzo colaborativo de
  recopilacion entre distintas fuentes (principalmente gobiernos
  locales). Nuevamente puede haber sesgos importanes.
\item
  En carácter insular de las Islas Canarias y de las Islas baleares no
  se ha tenido en cuenta.
\item
  Baleares se ha linkado artificialmente con 3/4 provincias y debería de
  haberse dejado aislada para respetar su carácter insular al definir la
  matriz W.
\item
  La incidencia depende de un estado inicial. Al inicio del estudio
  había provincias en las que la epidemia estaba muy desarrollada
  (Madrid/Alava) mientras que en otras apenas habías casos. Este hecho
  no ha sido considerado en el modelo. Decretar el confinamiento debe
  tener distintos impactos entre provincias. QUIZAR METER UNA VARAIBLE
  DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON
  MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
\item
  No se ha controlado por el sistema sanitario de cada provincia. Uno de
  los principales focos de contagio han sido los hospitales y los
  centros de salud. En aquellas provincias donde se ha promovido el
  mensaje ``NO IR AL MEDICO'' han presentado menor incidencia.
\item
  HAY QUE INCLUIR INTERVALOS DE CONFIANZA EN LOS GRÁFICOS QUE HACEMOS DE
  LOS COEFICIENTES DEL MODELO
\item
  ¿hasta que punto las variables de control no recojen también factores
  climáticos? por ejemplo, la gente jóven vive en el sur de España que
  ha tenido menos incidencia.
\item
  TITULO: A spatial analysis of the environmental correlates ofCOVID-19
  incidence in Spain ¿during the lookdown?
\item
  Las estaciones meteorológicas para la obtencion de datos climáticos
  has sido elegidas aleatoriamente (una para cada provincia). otra
  seleccion puede dar otros resultados. AQUI SE PODRÍA HACER EL ESFUERZO
  DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR
  PROVINCIA
\end{itemize}

*IDEM para la humedad. idem para sunshine

@@

\leavevmode\hypertarget{ref-goerlich2013population}{}%
Goerlich, F.J., Cantarino, I., 2013. A population density grid for
spain. International Journal of Geographical Information Science 27,
2247--2263.

@@

\leavevmode\hypertarget{ref-wang2020high}{}%
Wang, J., Tang, K., Feng, K., Lv, W., 2020. High temperature and high
humidity reduce the transmission of covid-19. Available at SSRN 3551767.
